---
title: "STAT432_HW8"
author: "Taiga Hasegawa(taigah2)"
date: "2019/3/26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Question1

Following logstic regression was done by built-in function. 
```{r}
    library(ElemStatLearn)
    data(SAheart)
    
    heart = SAheart
    heart$famhist = as.numeric(heart$famhist)-1
    n = nrow(heart)
    p = ncol(heart)
    
    heart.full = glm(chd~., data=heart, family=binomial)
    round(summary(heart.full)$coef, dig=3)
    
    # fitted value 
    yhat = (heart.full$fitted.values>0.5)
    table(yhat, SAheart$chd)
```

I'm going to replicate the above summary matrix using my own code.
```{r}
    # Gradient
    my.gradient <- function(b, x, y)
    {
        bm = as.matrix(b) 
        expxb =  exp(x %*% bm)
        return(t(x) %*% (y - expxb/(1+expxb)))
    }

    # Hessian
    my.hessian <- function(b, x, y)
    {
    	bm = as.matrix(b) 
    	expxb =  exp(x %*% bm)
    	x1 = sweep(x, 1, expxb/(1+expxb)^2, "*")
    	return(-t(x) %*% x1)
    }

    #Newton–Raphson
    my.logistic <- function(b, x, y, tol = 1e-10, maxitr = 30, gr, hess, verbose = FALSE)
    {
        b_new = b
        
        for (j in 1:maxitr) # turns out you don't really need many iterations
        {
        	b_old = b_new
        	b_new = b_old - solve(hess(b_old, x, y)) %*% gr(b_old, x, y)
        	
        	if (verbose)
        	{
        	    cat(paste("at iteration ", j,", current beta is \n", sep = ""))
        	    cat(round(b_new, 3))
        	    cat("\n")
        	}    
        	if (sum(abs(b_old - b_new)) < tol) break;
        }
        return(b_new)
    }
```

```{r}
    x = as.matrix(cbind("intercept" = 1, heart[, 1:9]))
    y = as.matrix(heart[,10])
    
    # set up an initial value
    b = rep(0, ncol(x))
    
    mybeta = my.logistic(b, x, y, tol = 1e-10, maxitr = 20, 
                         gr = my.gradient, hess = my.hessian, verbose = FALSE)
    mysd = sqrt(diag(solve(-my.hessian(mybeta, x, y))))  
    
    # my summary matrix
    round(data.frame("beta" = mybeta, "sd" = mysd, "z" = mybeta/mysd, 
    	  "pvalue" = 2*(1-pnorm(abs(mybeta/mysd)))), dig=3)
    
    # fitted value 
    yhat = (exp(x%*%mybeta)/(1+exp(x%*%mybeta))>0.5)
    table(yhat, SAheart$chd)
```
I got the exactly the same result with the one we found by built-in  function.

```{r eval=FALSE}
    # set up an initial value
    b = rep(1, ncol(x))

    mybeta = my.logistic(b, x, y, tol = 1e-10, maxitr = 20, 
                         gr = my.gradient, hess = my.hessian, verbose = FALSE)
    
    mysd = sqrt(diag(solve(-my.hessian(mybeta, x, y))))    
    
    # my summary matrix
    round(data.frame("beta" = mybeta, "sd" = mysd, "z" = mybeta/mysd, 
    	  "pvalue" = 2*(1-pnorm(abs(mybeta/mysd)))), dig=3)
```
I got the error because value in hessian matrix became too small to inverse. 

##Question2

a)

```{r}
findRows <- function(zip, n) {
# Find n (random) rows with zip representing 0,1,2,...,9 
res <- vector(length=10, mode="list")
names(res) <- 0:9
ind <- zip[,1]
for (j in 0:9) {
res[[j+1]] <- sample( which(ind==j), n ) }
return(res)
}
set.seed(1)
# find 100 samples for each digit for both the training and testing data
train.id <- findRows(zip.train, 100) 
train.id = unlist(train.id)
test.id <- findRows(zip.test, 100) 
test.id = unlist(test.id)
X = zip.train[train.id, -1] 
Y = zip.train[train.id, 1] 
dim(X)

 Xtest = zip.test[test.id, -1] 
 Ytest = zip.test[test.id, 1] 
 dim(Xtest)
```

```{r}
#prior 
pi_k=0.1

#centroid 
mu_k=sapply(0:9, function(x) apply(X[Y==x,],2,FUN=mean))

#covariance
Sigma=1/(1000-10)*(99*cov(X[Y==0,])+99*cov(X[Y==1,])+99*cov(X[Y==2,])+99*cov(X[Y==3,])+99*cov(X[Y==4,])+99*cov(X[Y==5,])+99*cov(X[Y==6,])+99*cov(X[Y==7,])+99*cov(X[Y==8,])+99*cov(X[Y==9,]))

#W_k
w_k=sapply(1:10, function(x) solve(Sigma)%*%mu_k[,x])

#b_k
b_k=sapply(1:10, function(x) -1/2*t(mu_k[,x])%*%solve(Sigma)%*%mu_k[,x]+log(pi_k))

#prediction
answer=sapply(1:1000,function(x) which.max(t(t(w_k)%*%t(Xtest)+b_k)[x,]))-1

#confusion matrix
table(Ytest, answer)

#prediction accuracy
mean(Ytest == answer)
```

b)

QDA does not work on this dataset. So I used one of the regularized approaches provided in the lecture note.
```{r}
library(mlbench)
library(caret)
library(glmnet)
library(klaR)
library(methods)
```

```{r warning=FALSE, error=FALSE}
set.seed(1337)
data=data.frame(zip.train[train.id, ])
data$X1=as.factor(data$X1)
cv_5_grid = trainControl(method = "cv", number = 5)
set.seed(1337)
fit_rda_grid = train(X1 ~ ., data = data, method = "rda", trControl = cv_5_grid)
fit_rda_grid
```

Next approach is that we reduced the dimension of the dataset first by PCA and then apply QDA.
```{r}
zip=rbind(X,Xtest)
pca=prcomp(zip)
plot(pca, type = "l", pch = 19, main = "Digits : PCA Variance")
pca=pca$x[,1:9]
X=pca[1:1000,]
Xtest=pca[1001:2000,]
dig.qda = qda(X, Y)
Ytest.pred=predict(dig.qda, Xtest)$class
table(Ytest, Ytest.pred)
mean(Ytest == Ytest.pred)
```
This approach is not as good as the previous one. 

##Question3

```{r}
#Prepare the dataset

y=c("No","No","Yes","Yes","Yes","No","Yes","No","Yes","Yes","Yes","Yes","Yes","No")

Xtrain=data.frame(
  "outlook"=c("Rainy","Rainy","Overcast","Sunny","Sunny","Sunny","Overcast","Rainy","Rainy","Sunny","Rainy","Overcast","Overcast","Sunny"),
  "temperature"=c("Hot","Hot","Hot","Mild","Cool","Cool","Cool","Mild","Cool","Mild","Mild","Mild","Hot","mild"),
  "humidity"=c("High","High","High","High","Normal","Normal","Normal","High","Normal","Normal","Normal","High","Normal","High"),
  "windy"=c("False","True","False","False","False","True","True","False","False","False","Ture","True","False","True"),
  "play_golf"=c("No","No","Yes","Yes","Yes","No","Yes","No","Yes","Yes","Yes","Yes","Yes","No")
  )

Xtest=c("Sunny","Hot","Normal","False")
```

```{r}
#Naive bayes
naivebayes=function(y,Xtrain,Xtest){
  prior_no=table(y)[1]/length(y)
  prior_yes=table(y)[2]/length(y)
  index=1
  condition_yes=rep(NA,length(Xtest))
  condition_no=rep(NA,length(Xtest))
  for(i in Xtest){
    X=Xtrain[Xtrain[,index]==i,]
    condition_yes[index]=sum(X[,ncol(X)]=="Yes")/table(y)[2]
    condition_no[index]=sum(X[,ncol(X)]=="No")/table(y)[1]
    index=index+1
  }
  posterior_yes=prod(condition_yes)*prior_yes
  posterior_no=prod(condition_no)*prior_no
  if(posterior_yes>posterior_no){
    cat("Yes")
  }else{cat("No")}
}

```

```{r}
naivebayes(y,Xtrain,Xtest)
```

