---
title: "STAT432_HW5"
author: "Taiga Hasegawa(taigah2)"
date: "2019/2/25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Question1

(a)$E[(Y-X\hat{\beta})^2]=E[(X\beta+e-X\hat{\beta})^2]=E[(X\beta+e-X\beta+X\beta-E[X\hat{\beta}]+E[X\hat{\beta}]-X\hat{\beta})^2]=E[(X\beta+e-X\beta)^2]+(X\beta-XE[\hat{\beta}])^2+E[(X\hat{\beta}-E(X\hat{\beta}))^2]=e^2+0+\sigma^2$

irreducible error: square error term(e^2). This cannot be reduced becuase we can't predict the e.
bias: 0 because OLS estimator is unbiased estimator. This evaluates how the average of our estimator deviates from the truth.
variance: $\sigma^2$ . This reflects the sensitivity of the function estimate f(x) to the training sample

(b) In ridge regression, the element of $\beta$ is shrinked by tuning parameter. $\beta^{ridge}=\alpha\beta^{OLS}$ where $\alpha$ is no more than 0. If we calculate the variance, it(tuning parameter>0) must be smaller than the variance of OLS estimator (tuning parameter=0) by $\alpha^2$, whereas it(tuning parametr>0) is biased because the OLS estimator is biased (tuning parameter=0). That's why as the tuning parameter increase, the vairance decreases and bias increases. 


##Question2
```{r}
data(Boston, package="MASS") 
head(Boston)
useLog = c(1,3,5,6,8,9,10,14) 
Boston[,useLog] = log(Boston[,useLog]) 
Boston[,2] = Boston[,2] / 10
Boston[,7] = Boston[,7]^2.5 / 10^4 
Boston[,11] = exp(0.4 * Boston[,11])/1000 
Boston[,12] = Boston[,12] / 100
Boston[,13] = sqrt(Boston[,13])
```

a)
```{r}
#fit a linear regression
fit=lm(medv~.,  data=Boston)
summary(fit)
```

(b)
```{r}
#Calculate the Mallow’s Cp statistic
n=nrow(Boston)
RSS=sum(fit$residuals^2)
p=ncol(Boston)
RSS+2*p*summary(fit)$sigma^2
```
```{r}
#Calculate the -2*loglikelihood
sigma2=RSS/n
loglikelihood= n*log(sigma2) + n + n*log(2*pi)+2
loglikelihood
```

```{r}
#Calculate the AIC
loglikelihood+2*p
```

```{r}
#Calculate the BIC
loglikelihood+log(n)*p
```

```{r}
#Best subset model selection 
library(leaps)
#maximum model size is 13
RSSleaps=regsubsets(as.matrix(Boston[,-14]),Boston[,14],nvmax=13)
sumleaps=summary(RSSleaps,matrix=T)
#calculate the AIC
msize=apply(sumleaps$which,1,sum)
AIC = n*log(sumleaps$rss/n) + 2*msize
#13 possible models
sumleaps$which
```
```{r}
#show which is the best model based on Cp 
which.min(sumleaps$cp)
```

```{r}
#show which is the best model based on AIC
which.min(AIC)
```
```{r}
#show which is the best model based on BIC
which.min(sumleaps$bic)
```
As a result, they all select the 9th model as the best one. 

This model has following variables: Intercept, chas, nox, rm, dis, rad, tax, ptratio, black and lstat.
```{r}
sumleaps$which[9,] 
```

##Question3
```{r}
library(MASS)
fit = lm.ridge(medv~., Boston, lambda=seq(0,100,by=0.1))
penaltylevel=which.min(fit$GCV)
penaltylevel
```

The best penelty level is `r penaltylevel` based on generalized cross-variation. 

The parameter estimates is like below.
```{r}
round(coef(fit)[which.min(fit$GCV), ], 4)
```

